---
title: "Using a random forest classifier to generate hair and eye colour predictions from SNP data"
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: 72
---

## Goal

This project evaluates the ability of a random-forest classifier to
correctly assign hair colour and eye colour labels to samples using
genotype data collected from 24 phenotype-informative SNP loci. While
the performance of phenotype classifiers is quite high in homogeneous
populations, the dataset used to test the RF classifier here includes
admixed individuals only. The exploration of admixed datasets is
essential to ensuring phenotype prediction can be used as a fair and
accurate tool in forensic contexts.

## Hair and Eye Colour Classes

There are three possible hair colours that can be assigned to a
sample: 1. Brown 2. Black 3. Blonde

There are three possible eye colours that can be assigned to a
sample: 1. Brown 2. Blue 3. Intermediate - This includes any non-brown
and non-blue eye colour, such as grey, green, and hazel.

The true hair colour and eye colour for each sample will be used to
determine if the classifier correctly identifies the group a sample
belongs to.

```{r include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r include=FALSE}
library(caret)
library(pROC) 
library(patchwork)
library(knitr)
library(tidyverse)
library(dplyr)
library(reshape2)
library(kableExtra)
```

```{r include=FALSE}
# Phenotype Data ----
phenotypes_df <- read.csv("PhenotypePredictions.tsv", sep="\t")

phenotypes_clean <- phenotypes_df %>%
  mutate(Eye.Self.ID = recode(Eye.Self.ID,
                              "Intermediate - Green" = "Intermediate",
                              "Intermediate - Hazel" = "Intermediate",
                              "brown" = "Brown"))
phenotypes_clean <- phenotypes_clean %>%
  mutate(Hair.Predicted = recode(Hair.Predicted,
                              "black" = "Black"))

phenotypes_clean <- phenotypes_clean %>%
  mutate(Eye.Predicted = recode(Eye.Predicted,
                                 "brown" = "Brown"))

lapply(phenotypes_clean, unique) # Check for improperly formatted values 

# Genotype Data ----
snp_df <- read.csv("snp_data.tsv", sep="\t")
snp_df$X <- sapply(snp_df$X, as.numeric) # X column has sample IDs
combined_df <- left_join(phenotypes_clean, snp_df, by = c("Sample" = "X")) # Combine using sample ID as key
combined_df_clean <- na.omit(combined_df) # Removed all NAs because only one sample is missing genotype data
SNP_cols = colnames(combined_df_clean[6:29])

# Transform data for model training ----
# Convert SNPs into levels
for (col in SNP_cols) {
  combined_df_clean[[col]] <- factor(combined_df_clean[[col]], levels = unique(combined_df_clean[[col]]))
}

# Convert phenotypes into factors 
hair_levels <- unique(combined_df_clean$Hair.Self.ID)
eye_levels <- unique(combined_df_clean$Eye.Self.ID)

combined_df_clean$Hair.Predicted <- factor(combined_df_clean$Hair.Predicted, levels = hair_levels)
combined_df_clean$Hair.Self.ID <- factor(combined_df_clean$Hair.Self.ID, levels = hair_levels)

combined_df_clean$Eye.Predicted <- factor(combined_df_clean$Eye.Predicted, levels = eye_levels)
combined_df_clean$Eye.Self.ID <- factor(combined_df_clean$Eye.Self.ID, levels = eye_levels)

# Train RF Classifier - Hair ----
set.seed(412)
train_index_hair <- createDataPartition(combined_df_clean$Hair.Self.ID, p=0.8, list=FALSE) # high p value to capture minority class in training data
train_data_hair <- combined_df_clean[train_index_hair, ]
test_data_hair <- combined_df_clean[-train_index_hair, ]
fit_control_hair <- trainControl(method = "cv", number = 5, sampling = "up") # Up sampled due to low counts in minority class (blonde) 

# Remove SNPs with only one level (aka all samples have the same genotype at this SNP)
valid_SNP_cols_hair <- SNP_cols[sapply(train_data_hair[ , SNP_cols], function(x) length(unique(x)) > 1)]
SNP_str_hair = paste(valid_SNP_cols_hair, collapse = " + ")

model_formula_hair <- paste("Hair.Self.ID ~", SNP_str_hair) # relationship between hair and SNPs
model_hair <- train(as.formula(model_formula_hair), 
                    data = train_data_hair,
                    method = "rf",
                    trControl = fit_control_hair)
pred_hair <- predict(model_hair, newdata = test_data_hair)
prob_hair <- predict(model_hair, newdata = test_data_hair, type = "prob")

# Train RF Classifier - Eyes ----
set.seed(412)
train_index_eye <- createDataPartition(combined_df_clean$Eye.Self.ID, p=0.8, list=FALSE) # High p value to capture minority class in training data 
train_data_eye <- combined_df_clean[train_index_eye, ]
test_data_eye <- combined_df_clean[-train_index_eye, ]
fit_control_eye <- trainControl(method = "cv", number = 5, sampling = "up") # Up sampled due to low counts in minority class (intermediate) 

# Remove SNPs with only one level
valid_SNP_cols_eye <- SNP_cols[sapply(train_data_eye[ , SNP_cols], function(x) length(unique(na.omit(x))) > 1)]
SNP_str_eye = paste(valid_SNP_cols_eye, collapse = " + ")

model_formula_eye <- paste("Eye.Self.ID ~", SNP_str_eye) # Relationship between eye colour and SNPs
model_eye <- train(as.formula(model_formula_eye), 
                    data = train_data_eye,
                    method = "rf",
                    trControl = fit_control_eye)

pred_eye <- predict(model_eye, newdata = test_data_eye)
prob_eye <- predict(model_eye, newdata = test_data_eye, type = "prob")
```

```{r}
# Create confusion matrices
conf_matrix_hair <- confusionMatrix(pred_hair, test_data_hair$Hair.Self.ID)
conf_matrix_eye <- confusionMatrix(pred_eye, test_data_eye$Eye.Self.ID)
```

## Random Forest Classifier Performance

The performance of hair and eye colour predictions were assessed
separately.

### Hair Colour

Due to the small number of blonde samples, the metrics in Table 2 for
the blonde class are heavily skewed.

```{r}
# Format confusion matrix table ----
conf_table_hair <- conf_matrix_hair$table
conf_df_hair <- as.data.frame.matrix(conf_table_hair)
kable(conf_df_hair, caption = "Hair Colour Confusion Matrix", booktabs = TRUE, escape = FALSE) %>%
  kable_styling(latex_options = c("hold_position")) %>%
  add_header_above(c("True" = 1, "Predicted" = 3))

# Format classifier metrics table ----
hair_metrics <- round(conf_matrix_hair$byClass[, 1:7], 3)
kable(hair_metrics, caption = "Hair Colour Classifier Metrics")

# ROC AUC Curves ----
true_labels_hair <- test_data_hair$Hair.Self.ID 
actual_mat_hair <- model.matrix(~ true_labels_hair - 1) 
colnames(actual_mat_hair) <- levels(true_labels_hair)
```

```{r roc-plots}
# ROC plots

roc_list_hair <-list() 
auc_list_hair <- c()
roc_list_plot_hair <- list()
roc_objects <- list()
auc_values <- list()

# Create plot for each class
for (class in levels(true_labels_hair)) {
  roc_obj <- suppressMessages(roc(actual_mat_hair[, class], prob_hair[, class]) )
  auc_val <- auc(roc_obj) 
  roc_plot <- ggroc(roc_obj) +
    ggtitle(paste("Hair -", class, "\nAUC:",round(auc_val, 3))) +
    theme_minimal() +
    theme(plot.title = element_text(size = 10)) +
    xlab("False Positive Rate") +
    ylab("True Positive Rate")
  roc_list_hair[[class]] <- roc_obj 
  auc_list_hair[class] <- auc_val 
  roc_list_plot_hair[[class]] <- roc_plot
}
roc_list_plot_hair[["Blonde"]] | roc_list_plot_hair[["Brown"]] | roc_list_plot_hair[["Black"]]

kable(as.data.frame(auc_list_hair), col.names = c("AUC"), caption ="Hair Colour - AUC per Class")

```

### Eye Colour
Due to a low number of samples in the intermediate class, the performance metrics for this class shown in Table 5 are heavily skewed.

```{r}
# Format confusion matrix table ----
conf_table_eye <- conf_matrix_eye$table
conf_df_eye <- as.data.frame.matrix(conf_table_eye)
kable(conf_df_eye, caption = "Eye Colour Confusion Matrix", booktabs = TRUE, escape = FALSE) %>%
  kable_styling(latex_options = c("hold_position")) %>%
  add_header_above(c("True" = 1, "Predicted" = 3))

# Format classifier metrics table ----
eye_metrics <- round(conf_matrix_eye$byClass[, 1:7], 3)
kable(eye_metrics, caption = "Eye Colour Classifier Metrics")
```

```{r include=FALSE}
# One-hot encode true labels

true_labels_eye <- test_data_eye$Eye.Self.ID 
actual_mat_eye <- model.matrix(~ true_labels_eye - 1) 
colnames(actual_mat_eye) <- levels(true_labels_eye)
```

```{r roc-plots-eye}
# ROC plots

roc_list_eye <- list()
auc_list_eye <- c()
roc_list_plot <- list()
roc_objects <- list()
auc_values <- list()

for (class in levels(true_labels_eye)) {
  roc_obj <- suppressMessages(roc(actual_mat_eye[, class], prob_eye[, class]))
  auc_val <- auc(roc_obj) 
  roc_plot <- ggroc(roc_obj) +
    ggtitle(paste("Eye -", class, "AUC:", round(auc_val, 3))) +
    theme_minimal() +
    theme(plot.title = element_text(size = 10)) +
    xlab("False Positive Rate") +
    ylab("True Positive Rate")
  roc_list_eye[[class]] <- roc_obj 
  auc_list_eye[class] <- auc_val 
  roc_list_plot[[class]] <- roc_plot
}

roc_list_plot[["Blue"]] | roc_list_plot[["Brown"]] | roc_list_plot[["Intermediate"]]


kable(as.data.frame(auc_list_eye),
  col.names = c("AUC"), caption = "Eye Colour - AUC per Class")
```

## Conclusions 
The interpretation of these results is limited due to the small sample size overall. The current metrics show the RF model is most effective at classifying brown eyes and black hair. If this trend holds true in a larger sample, this would indicate that the 24 SNPs utilized in this study are most informative for brown eyes and black hair, but less informative for other hair and eye colour classes. The addition of more SNPs to phenotyping panels may assist in more effective classification of blonde and brown hair, and intermediate and blue eyes. 